Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities

### 2022-04-27 paper/manuscript.Rmd
Cleanup, PNAS formatting

---
title: Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities
authors: 
  - name: "Kari E. A. Norman"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: kari.norman@berkeley.edu
  - name: "Perry de Valpine"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: pdevalpine@berkeley.edu
  - name: "Carl Boettiger"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
 Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing trait data representative of a species' functional role. We found no overall trend in any functional diversity metric, despite similar species-based patterns of constant richness with increasing turnover. The lack of trend held even after correcting for changes in species richness and at the study-level, where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric. Results give evidence that across a variety of taxa, climates, and biomes, these selected functional characteristics are maintained even in the face of significant environmental and community change. We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  rticles::arxiv_article:
    keep_tex: true
#  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
keywords:
  - biodiversity change
  - functional traits
  - global change
  - timeseries

@@

It is critical to understand how ecological communities are responding to the ongoing biodiversity crisis. Previous broad-scale assessments of community change through time have focused on metrics of species identity, which do not account for differences between species. Functional trait approaches offer a powerful alternative to species-based approaches by directly assessing changes in the species characteristics that define their ecological role. Here, we pair vertebrate community timeseries with functional trait data to take a first look at temporal trends in functional composition across biomes, realms, and taxa. We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects of functional structure. We also identified key methodological holes that may be obscuring a potential underlying trend.

@@

```{r echo = FALSE, results = "asis"}
tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
       Timeseries_Count = c(48, 2380, 11),
       Species_Count = c(184, 700, 184),
       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
) %>%
  mutate_all(linebreak, align = "l") %>% 
  kableExtra::kable(format = "latex", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
               caption = "Summary of the data in the final trait database.") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
   #kableExtra::column_spec(2, width = "5cm") #%>%
  # kableExtra::column_spec(4, width = "10cm")
```

We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
These databases include species-level means for traits that partially represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
For the full list of traits included in the analysis for each taxon see Table 1.
Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.

In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the *taxadb* R package [@norman2020; @rcoreteam2021].
If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
For trait data, traits for all species of the same identifier were averaged.
Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.

Our final dataset included 2,443 time series from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig 1).
The earliest sample was in 1923 and the most recent was in 2016.
Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table 1).
Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include data while acknowledging that general inference for amphibians as a clade is not possible with the time series available.
For a full breakdown of studies and their characteristics, see the supplement.

```{r taxa-map, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
```

We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
All timeseries with data for a given trait were included in the model.
Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
We fit 24 individual *CWM* models.

```{r timeseries-plot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
knitr::include_graphics(here::here("figures/3met_long.jpeg"))
```

@@

We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig 2).
We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table 2).
Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
However, with only two global studies and two freshwater studies these results are not general trends.
We found similar results for the *CWM* models, with a significant trend for only one trait, percentage of aerial foraging utilization, which showed a significant positive trend.

```{r echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything()) %>%
  filter(group != "Residual" | is.na(group)) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
         metric = ifelse(metric == "S", "Species Richness", metric),
         effect = ifelse(effect == "ran_pars", "random", effect),
         group = case_when(
           group == "rarefyID:study_id" ~ "timeseries within study",
           group == "study_id" ~ "study",
           group == "Residual" ~ "residual",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "SD Intercept",
           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
           term == "sd__year_scaled" ~ "SD Year",
           term == "sd__Observation" ~ "SD Observation",
           TRUE ~ term
         )) %>%
  mutate(across(c(estimate, std.error, p.value), round, 2),
         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
  dplyr::rename(grouping = group) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics. Additional model estimates, including CWM models, can be found in the supplement.") %>%
  column_spec(3, width = "2cm") %>%
  collapse_rows(columns = 1:3, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("scale_down", "hold_position"))
```

At the study level, 4 studies experienced a significant trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table 3).
Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
However, the majority of the functional richness trends were negative, indicating that when change is occurring it is more often a loss.
Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
Rather than interpreting changes in specific studies, we present these results as a general picture of how few studies experienced a significant trend and highlight that even that count is likely an overestimate.

```{r echo = FALSE}
library(tidyverse)

@@

While we did not directly measure changes in rare species, our results further contradict likely scenarios of loss predicted due to rare species extinction.
Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species can eventually no longer be replaced by functionally similar species [@leitão2016].
Observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.
Still, for many timeseries we likely did not have a large enough time window to capture community and species pool impoverishment due to extinction.

@@

Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
However, a number of gaps in best practices may be obscuring a true underlying trend.
First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
Most time series span only a few years (Figure 1) and may not provide the statistical power necessary to detect trends.
The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021], particularly scenarios of conversion to urban, human-dominated landscapes.
While evidence from other work shows even disturbed communities can maintain functional structure [@edwards2013; @matuoka2020], these results should not be interpreted as evidence of low functional impact in areas of heavy human disturbance.

@@

>>>>

---
title: Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities
authors: 
  - name: "Kari E. A. Norman"
    affiliation: Université de Montréal
    location: Montréal, Quebec
    email: kari.norman@berkeley.edu
  - name: "Perry de Valpine"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: pdevalpine@berkeley.edu
  - name: "Carl Boettiger"
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
 Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time. Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes. Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing trait data representative of a species' functional role. We found no overall trend in any functional diversity metric, despite similar species-based patterns of constant richness with increasing turnover. The lack of trend held even after correcting for changes in species richness and at the study-level, where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric. Results give evidence that across a variety of taxa, climates, and biomes, these selected functional characteristics are maintained even in the face of significant environmental and community change. We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.

journal: "Ecology Letters"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: pnas.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  rticles::arxiv_article:
    keep_tex: true
#  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
keywords:
  - biodiversity change
  - functional traits
  - global change
  - time series

@@

It is critical to understand how ecological communities are responding to the ongoing biodiversity crisis. Previous broad-scale assessments of community change through time have focused on metrics of species identity, which do not account for differences between species. Functional trait approaches offer a powerful alternative to species-based approaches by directly assessing changes in the species characteristics that define their ecological role. Here, we pair vertebrate community time series with functional trait data to take a first look at temporal trends in functional composition across biomes, realms, and taxa. We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects of functional structure. We also identified key methodological holes that may be obscuring a potential underlying trend.

@@

@@

We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig \ref{fig:timeseriesPlot}).
We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
However, with only two global studies and two freshwater studies these results are not general trends.
We found similar results for the *CWM* models, with a significant trend for only one trait, percentage of aerial foraging utilization, which showed a significant positive trend.

```{r resultsTab, echo=FALSE}
read_csv(here::here("figures/man_model_table.csv")) %>%
  select(metric, everything()) %>%
  filter(group != "Residual" | is.na(group)) %>%
  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
         metric = ifelse(metric == "S", "Species Richness", metric),
         effect = ifelse(effect == "ran_pars", "random", effect),
         group = case_when(
           group == "rarefyID:study_id" ~ "time series within study",
           group == "study_id" ~ "study",
           group == "Residual" ~ "residual",
           TRUE ~ group),
         term = case_when(
           term == "(Intercept)" ~ "Intercept",
           term == "year_scaled" ~ "Year",
           term == "sd__(Intercept)" ~ "SD Intercept",
           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
           term == "sd__year_scaled" ~ "SD Year",
           term == "sd__Observation" ~ "SD Observation",
           TRUE ~ term
         )) %>%
  mutate(across(c(estimate, std.error, p.value), round, 2),
         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
  dplyr::rename(grouping = group) %>%
  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics. Additional model estimates, including CWM models, can be found in the supplement.") %>%
  column_spec(3, width = "2cm") %>%
  collapse_rows(columns = 1:3, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("scale_down", "hold_position"))
```

At the study level, 4 studies experienced a significant trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table \ref{tab:trendTab}).
Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
However, the majority of the functional richness trends were negative, indicating that when change is occurring it is more often a loss.
Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
Rather than interpreting changes in specific studies, we present these results as a general picture of how few studies experienced a significant trend and highlight that even that count is likely an overestimate.

```{r trendTab, echo = FALSE}
library(tidyverse)

@@

While we did not directly measure changes in rare species, our results further contradict likely scenarios of loss predicted due to rare species extinction.
Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species can eventually no longer be replaced by functionally similar species [@leitão2016].
Observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.
Still, for many time series we likely did not have a large enough time window to capture community and species pool impoverishment due to extinction.

@@

Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
However, a number of gaps in best practices may be obscuring a true underlying trend.
First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
Most time series span only a few years (Fig \ref{fig:taxaMap}) and may not provide the statistical power necessary to detect trends.
The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021], particularly scenarios of conversion to urban, human-dominated landscapes.
While evidence from other work shows even disturbed communities can maintain functional structure [@edwards2013; @matuoka2020], these results should not be interpreted as evidence of low functional impact in areas of heavy human disturbance.

@@

```{r traitSources, echo = FALSE, results = "asis"}
tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
       Timeseries_Count = c(48, 2380, 11),
       Species_Count = c(184, 700, 184),
       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
) %>%
  mutate_all(linebreak, align = "l") %>% 
  kableExtra::kable(format = "latex", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
               caption = "Summary of the data in the final trait database.") %>%
  kable_classic() %>% 
  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
   #kableExtra::column_spec(2, width = "5cm") #%>%
  # kableExtra::column_spec(4, width = "10cm")
```

We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
These databases include species-level means for traits that partially represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
For the full list of traits included in the analysis for each taxon see Table \ref{tab:traitSources}.
Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.

In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the `taxadb` R package [@norman2020; @rcoreteam2021].
If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
For trait data, traits for all species of the same identifier were averaged.
Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.

Our final dataset included 2,443 time series from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig \ref{fig:taxaMap}).
The earliest sample was in 1923 and the most recent was in 2016.
Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table \ref{tab:traitSources}).
Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include data while acknowledging that general inference for amphibians as a clade is not possible with the time series available.
For a full breakdown of studies and their characteristics, see the supplement.

```{r taxaMap, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
```

We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
All time series with data for a given trait were included in the model.
Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
We fit 24 individual *CWM* models.

```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
knitr::include_graphics(here::here("figures/3met_long.jpeg"))
```
